Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-04-30 analysis/paper/paper.Rmd
minor edits to intro, methods, results

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009]. In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020]. Impacts on wildlife persist after roads are no longer driveable as they to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021]. Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.

@@

We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000]. Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing. However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.  We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster. If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context. Both Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al. [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water. We also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.

@@

Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. We assumed everything in the digital road atlas was built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a construction cost of 0. We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells. In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells. In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure @\ref(fig:study-area)b). This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method. The road construction cost and elevation rasters consist of 1.7 million    pixels with a resolution of 1 ha. We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters. We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.

@@

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006]. The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl, Supplement X). Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004]. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.

@@

Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks. As an example, distance to roads within cutblocks is the average distance from non-road pixels within cutblocks to the nearest roads. We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and supplemental figure with old results). We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)). The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density. All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projects of road presence and distance to road.

Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads. Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)). The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c). Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms. Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g. \@ref(fig:proj-map-fig)b). We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement B).

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure C.1). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure C.1).

@@

bench_res %>% filter(cutblocks_real == "revelstoke_real", sampleDens == 1e-05,
                     sampleType == "regular") %>% 
  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
               values_to = "values") %>% 
  ggplot(aes(y = values, x = toupper(method), fill = as.factor(resolution)))+
  geom_col(position = position_dodge())+
  scale_fill_discrete(type = met.brewer(pal_nm, 3))+
  facet_wrap(~variable, nrow = 2, scales = "free", 
             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
                                              Peak_RAM_Used_MiB = "Peak RAM used (MiB)"))) +
  labs(y = NULL, x = "Projection method", fill = "Resolution")

@@

Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement D). All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e. not NA), so options such as increasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements. Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g. only target the centroid of each cutblock). Hardy comparison....

>>>>

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009]. In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020]. Impacts on wildlife persist after roads are no longer driveable as they continue to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021]. Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.

@@

We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000]. Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing. However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.  We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster. If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context. Both Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al. [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water. Here we assess this approach and also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.

@@

Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. We assumed everything in the digital road atlas was built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections. We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks. Results including all cutblocks are also presented.

Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a construction cost of 0. We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells. In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells, and we refer to it as the “simple cost” method. In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b). This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method. The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha. We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters. We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.

@@

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006]. The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A). Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004]. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.

@@

Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks. As an example, distance to roads within cutblocks is the average distance from non-road pixels within cutblocks to the nearest roads. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure B.1). We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)). The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density. All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.

Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads. Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)). The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c). Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms. Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g. \@ref(fig:proj-map-fig)b). We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement C).

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure D.1).

@@

# Hardy time measured in QGIS with 100m raster
hardy_time <- 2*60+34

@@

Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement E). All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e. not NA), so options such as decreasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements. Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g. only target the centroid of each cutblock). The Hardy method, run in QGIS, was an order of magnitude faster than any of our methods run in R.
